---
title: "Market Basket Analysis"
author: "Njabulo Hlabangana"
date: '2022-07-23'
output: 
  html_document:
    theme: cerulean
---


```{r setup, include=FALSE,warning=FALSE,message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## What is market basket analysis (MBA)?
This is a tried and tested technique that is used to ascertain association rules in a data set. It is mainly used by retailers to determine which items are often bought together as a 'basket'. The technique,essentially, helps retailers determine how likely it is for a shopper to buy a particular product given that they have already bought a particular set of product(s).

Brick and motor sellers can use MBA to allocate shelve space - trying as much as possible to ensure that products that tend to be bought together are kept as close to each other as possible. This helps make shopping convenient for customers and  can potentially boost sells. Online retailers can use MBA to come up with product recommendations based on the products they already have in their shopping carts in the process of shopping.

## Dataset Used In The Analysis

The data was obtained from the Meta Marketing Analytics Course that I did on Coursera. It was used for different kinds of analysis there and I figured it could be great to go a step further and perform MBA on it. 

The data contains transactions on a small company - Inu & Neko that specializes in selling pet (cats and dogs) food, toys, bedding and other treats. 




```{r message=FALSE, warning= FALSE}
#loading packages to be used in the analysis
library(tidyverse)
library(arulesViz)

```

## Loading the data

A quick look at at the data shows that it is arranged in a **single** format. This is a format in which a transaction ID is assigned to a basket (collection of items bought together) and every item in that basket is identified using that transaction ID. The transaction ID are arranged as rows. This means that a basket with three items will have three rows in that data set, each row with the same ID but different items.

```{r, echo=TRUE}
setwd('/Users/hlaba/Documents/Custom Office Templates/BA') # setting the working directory to the folder that contains the data set
dat <- read.csv('inu_neko_orderline_clean.csv') # reading in the csv file
head(dat)
```

## Needed Columns

Only two columns will be needed for the analysis viz, transaction ID and product name. The code below makes a dataframe out of those two columns. Since the function used to read transactions cannot read dataframes, the dataframe will be converted to a csv file and saved in the same working directory.

```{r}
transactions <- dat %>% select(as.factor('trans_id'),as.factor('prod_title'))
head(transactions)
write.csv(transactions,'/Users/hlaba/Documents/Custom Office Templates/BA/transactions.csv',row.names = F)
```
## Reading in the transaction

The data frame has to be converted into a format that can the handled by the arules package as shown below.The inspect command is used together with the head command to show the selected number of baskets and their respective transaction IDs.

```{r, echo=TRUE}
purchases <- read.transactions('transactions.csv',format = 'single', sep = ",",cols = c(1,2),header = T) # reading the two columns of the single format dataframe.
inspect(head(purchases,5))
```
## Some exploratory data analysis
To get a better view of the data, we will use the summary command. This shows that:
  there is a total of 28022 transactions involving 21 different items offered by the Inu & Neko.      
  The most common basket size is actually 1 with 20808 transactions.
  Reddy Beddy is the most frequently purchased product. This is also shown clearly in the frequency plot below

```{r}
summary(purchases)
itemFrequencyPlot(purchases,topN = 21, type = 'relative')
```


## Finding the association rules

The association rules are found using the appriori algorithm as shown below. **Support** refers to the count of a particular basket as a proportion of the         total number of unique baskets. The entered value for support is the threshold needed for the basket to be included. A higher value results in a fewer baskets. In this case the value is 0.01 %. This was chosen because backet sizes greater than one constitute a small proportion of the transactions - 25.74%. The **minlen** and **maxlen** arguments stipulate the minimum and and maximum basket size respectively. The **confidence** value is the minimum posterior probability for a particular item to be in bought given a particular basket. More on this in the interpretation of output. When **verbose** is set to TRUE the algorithm will display the steps it goes through on the console. I do not think that is necessary so I set it to FALSE.

```{r}
rules <- apriori(purchases, parameter = list(support = 0.0001,minlen = 2,maxlen = 6, confidence = 0.5),control = list(verbose = F))

```
## A quick look at the rules
The mined rules are ordered by **lift**. This is essentially the number of times the item frequency(prior probability) is increased given that a particular set of items in the left hand side(LHS) have already been bought or are in the cart. Purrfect Puree has the highest lift - 11.78. This means that a customer who has bought or wants to but the LHS items (Feline Fix Mix,Kitty Climber,Purr Mix) is 11.78 times more likely to buy Purrfect Puree than a customer who does not have those items in the basket or has no intention of buying them. The confidence for the same basket is 0.75. This essentially means that about 75% of the baskets with the LHS items (Feline Fix Mix,Kitty Climber,Purr Mix) will also contain the Purrfect Puree. It would therefore make more sense to allocate shelf space in such a way that these items are close to each other. If the shopping is done online, then Perrfect Puree should be recommended to a customer who has bought Feline Fix Mix,Kitty Climber and Purr Mix or has them in their cart.

```{r}
inspect(head(sort(rules, by ="lift"),5))
```
## Visualization of Results
The following graphs help to visualize the mined rules. The first is called a grouped graph and it arranges rules by order of lift and support. AS expected the rule with the highest lift as discussed above appears at the top. The number of rules in the graph was reduced to 5 to avoid clutter.
```{r}
plot(unique(rules), method = 'grouped', k = 5)
plot(rules,method = 'graph')
```

## Conclusion

Based on its structure, the dataset lends itself well to analysis using market basket analysis. However, the fact that the majority (about 76%) of the baskets are one-item baskets makes the sample size for mining association rules very small in ways that can affect the reliability of the results. The principles and methods do, however, remain the same and applicable to bigger datasets.  


